% !TEX root = EnergyTrading_ss13UDE.tex
\section{Energy Derivatives}
\subsection{Basic Pricing Relations for Forwards/Futures}

Storage, Inventory and Convenience Yield
	The theory of storage aims to explain the differences between spot
	and Futures (Forward) prices by analyzing why agents hold inventories.
	
	Inventories allow to meet unexpected demand, avoid the cost of frequent revisions in
	the production schedule and eliminate manufacturing disruption.
	
	This motivates the concept of convenience yield as a benefit, that accrues to the
	owner of the physical commodity but not to the holder of a forward contract.
	
	Thus the convenience yield is comparable to the dividend yield for stocks.
	
	A modern view is to view storage (inventory) as a timing option, that
	allows to put the commodity to the market when prices are high
	and hold it when the prices are low.


Spot-Forward Relationship in Commodity Markets
	Under the no-arbitrage assumption we have
	\begin{equation}\label{SF-rel}
	F(t,T)=S(t)e^{(r-y)(T-t)}
	\end{equation}
	where $r$ is the interest rate at time $t$ for maturity $T$ and $y$ is the convenience yield

	Observe that (\ref{SF-rel}) implies
		spot and forward are redundant (one can replace the other) and form a linear relationship (unlike options)
	
		with two forward prices we can derive the value of $S(t)$ and $y$
		
		knowledge of $S(t)$ and $y$ allows us to construct the whole forward curve
		
		for $r-y <0$ we have backwardation; for $r-y>0$ we have contango.


Spot-Forward Relationship: Classical theory
	In a stochastic model we use
		$$
		F(t,T)=\EX_{\Q}(S(T)|\F_t)
		$$
	where $\F_t$ is the accumulated available market information (in most models the information generated by the spot price).
	$\Q$ is a risk-neutral probability

	discounted spot price is a $\Q$-martingale
	
	fixed by calibration to market prices or a market price of risk argument


Futures Prices and Expectation of Future Spot Prices
	The rational expectation hypothesis (REH) states that the current futures price $f(t,T)$ for a commodity with
	delivery a time $T>t$ is the best estimator for the price $S(T)$ of the commodity.
	In mathematical terms
		\begin{equation}\label{REH}
		f(t,T) = \EX[S(T) |\F_t].
		\end{equation}
	where $\F_t$ represents the information available at time $t$. The REH has been statistically
	tested in many studies for a wide range of commodities.

	When equality in (\ref{REH}) does not hold futures prices are biased estimators of
	future spot prices. If
		[$>$] holds, then $f(t,T)$ is an up-ward biased estimate, then risk-aversion
		among market participants is such that buyers are willing to pay more than the expected
		spot price in order to secure access to the commodity at time $T$ (political unrest);
		
		[$<$] holds, then $f(t,T)$ is an down-ward biased estimate, this may reflect a
		perception of excess supply in the future.


Market Risk Premium
	The \emph{market risk premium} or \emph{forward bias} $\pi (t,T)$
	relates forward and expected spot prices.
	
	It is defined as the difference, calculated at time $t$, between
	the forward $F(t,T)$ at time $t$ with delivery at $T$ and expected
	spot price:
		\begin{equation}\label{forward risk premium}
		\pi(t,T)= F(t,T)-\EX^\prb[S(T)|\mathcal F_t].
		\end{equation}
	Here $\EX^\prb$ is the expectation operator, under the
	historical measure $\prb$, with information up until time $t$ and
	$S(T)$ is the spot price at time $T$.


\subsection{HJM-type models}
Heath-Jarrow-Morton (HJM) model}
	The Heath-Jarrow-Morton model uses the entire forward rate curve as
	(infinite-dimensional) state variable. The dynamics of the forward rates $f(t,T)$ are {\it exogenously} given by
	$$
	%\label{forward rate dynamics}
	df(t,T) = \alpha(t,T) dt + \s(t,T) dW(t).
	$$
	For any fixed maturity $T$, the
	initial condition of the stochastic differential equation
	is determined by the current value
	of the empirical (observed) forward rate for the future date $T$
	which prevails at time $0$.


One-Factor GBM Specification
	Here the volatility is
		$$
		\sigma_1(t,T)=e^{-\kappa (T-t)}\sigma
		$$
	and
		$$
		dF(t,T)=F(t,T)\sigma_1(t,T)dW(t)
		$$
		

Two-Factor GBM Specification
	Here the volatilities are
		$$
		\sigma_1(t,T)=e^{-\kappa (T-t)}\sigma_1 \; \mbox{ and } \; \sigma_2>0
		$$
	and
		$$
		\frac{dF(t,T)}{F(t,T)}=\sigma_1(t,T)dW_1(t)+\sigma_2dW_2(t)
		$$


\subsection{A Market Model}
Modelling Approach
	We use the HJM-framework to model the forward dynamics directly.
	
	We distinguish between forward contracts with a fixed time delivery 
	and forward contracts with a delivery period, called \emph{swaps}.
	
	Since the HJM-framework cannot be applied to the swap dynamics literally, 
	we differentiate between the \emph{decomposable} swaps and the \emph{atomic} 
	swaps and apply the framework to the atomic swaps.

	The dynamics of a decomposable swap with delivery period $[T_1,T_{N}]$ can than be obtained from $N-1$ atomic swaps by
	\begin{align}
	F(t,T_1,T_N)=\sum\limits_{i=1}^{N-1} \frac{T_{i+1}-T_i}{T_N-T_1} F(t,T_i,T_{i+1})\label{eqn: decomposbale swap}
	\end{align}

	We discuss several lognormal dynamics of the swap price,
		\begin{equation}
		dF(t,T_1,T_2)=\Sigma(t,T_1,T_2)F(t,T_1,T_2)\, dW(t). \label{eqn: lognormal dynamics}
		\end{equation}
	The only parameter in this model is the volatility function $\Sigma$ which has to capture all movements of the swap price and especially the time to maturity effect.


Volatility Functions
	We assume that the swap price dynamics \emph{for all atomic swaps} is given by (\ref{eqn: lognormal dynamics})
	where $\Sigma(t,T_1,T_2)$ is a continuously differentiable and positive function.

	Starting out with a given volatility function for a fixed time forward contract we see that the volatility function $\Sigma$ for the swap contract is given by
	\begin{equation}
	\Sigma(t,T_1,T_2)=\int_{T_1}^{T_2} \hat{w}(u,T_1,T_2) \sigma(t,u) \, du. \label{eqn: swap volatility creation}
	\end{equation}


Schwartz
	For the volatility function of the forward we use
	\begin{equation}\label{vol-schwartz}
	\sigma(t,u)=a e^{-b(u-t)}
	\end{equation}
	where $a,b >0 $ are constant.

	The time to maturity effect is modeled by a negative exponential function.
		When the time to maturity tends to infinity the volatility function converges to zero.
		
		The exponential function causes that the volatility increases as the time to maturity decreases 
		which leads to an increased volatility when the contract approaches the maturity.

	Applying this forward volatility to (\ref{eqn: swap volatility creation}) the swap volatility is:
		\begin{align}
		\Sigma(t,T_1,T_2)&=a\,\varphi(T_1,T_2)
		\end{align}
	where
		\begin{align}
		\varphi(T_1,T_2)= \frac{e^{-b(T_1-t)}-e^{-b(T_2-t)}}{b(T_2-T_1)}
		\label{volatility function varphi}
		\end{align}
	The Black-76 specification of the forward volatility can be obtained if $\varphi(T_1,T_2) =1$, that is $b=0$
	in (\ref{vol-schwartz}).

	The associated swap price volatility is then given by $\Sigma(t,T_1,T_2)=a$.


Model Summary
A summary of the different models presented above is listed in Table \ref{tab: lognormal models}.
	\begin{table}[h]
	 \small \centering {\sf
	\begin{tabular}[t]{l|c}
					Model & $\Sigma(t,T_1,T_2)$ \\ %\toprule
				 Black-76 & $a$ \\
					Schwartz  &$a\, \varphi(T_1,T_2)$  \\
					Fackler and Tien  & $a(t)\, \varphi(T_1,T_2)$ \\
					Clewlow and Strickland & $a((1-c)\, \varphi(T_1,T_2)+c)$ \\
					Koekebakker and Lien  & $a(t)((1-c)\, \varphi(T_1,T_2)+c)$ \\
					Benth and Koekebakker& $\hat{a}\, \varphi(T_1,T_2) +a(t)$ \\

			\end{tabular}}
			\caption[Swap Volatility Models]
			{\small The associated swap volatility models generated by (\ref{eqn: swap volatility creation}) with $a \ge 0, b > 0$ and $0 \le c \le 1$ constants, $a(t)$ defined in (\ref{seasonality function}) and $\varphi(T_1,T_2)$ is given by (\ref{volatility function varphi}). }
			\label{tab: lognormal models}
	\end{table}

