% !TEX root = StructuringValuation_ws1415UDE.tex
\part{Forward Markets}
\section{Basic Definitions and Models}
\subsection{Basic Pricing Relations for Futures}

% frametitle
{Storage, Inventory and Convenience Yield}
\begin{itemize}
\item<1-> The theory of storage aims to explain the differences between spot
and Futures (Forward) prices by analyzing why agents hold inventories.
\item<2-> Inventories
allow to meet unexpected demand, avoid the cost of frequent revisions in
the production schedule and eliminate manufacturing disruption.
\item<3-> This
motivates the concept of convenience yield as a benefit, that accrues to the
owner of the physical commodity but not to the holder of a forward contract.
\item<4-> Thus the convenience yield is comparable to the dividend yield for stocks.
\item<5-> A modern view is to view storage (inventory) as a timing option, that
allows to put the commodity to the market when prices are high
and hold it when the prices are low.
\end{itemize}

% frametitle
{Spot-Forward Relationship in Commodity Markets }
Under the no-arbitrage assumption we have
\begin{equation}\label{SF-rel}
F(t,T)=S(t)e^{(r-y)(T-t)}
\end{equation}
where $r$ is the interest rate at time $t$ for maturity $T$ and $y$ is the convenience yield.

% frametitle
{Spot-Forward Relationship in Commodity Markets }

Observe that (\ref{SF-rel}) implies
\begin{itemize}
\item<1-> spot and forward are redundant (one can replace the other) and form a
linear relationship (unlike options)
\item<2-> with two forward prices we can derive the value of $S(t)$ and $y$
\item<3-> knowledge of $S(t)$ and $y$ allows us to construct the whole forward curve
\item<4-> for $r-y <0$ we have backwardation; for $r-y>0$ we have contango.
\end{itemize}

% frametitle
{Spot-Forward Relationship: Classical theory}
\begin{itemize}
\item<1-> In a stochastic model we use
$$
F(t,T)=\EX_{\Q}(S(T)|\F_t)
$$
where $\F_t$ is the accumulated available market information (in most models the information generated by the spot price).
\item<2-> $\Q$ is a risk-neutral probability
\begin{itemize}
\item discounted spot price is a $\Q$-martingale
\item fixed by calibration to market prices or a market price of risk argument
\end{itemize}
\end{itemize}

% frametitle
{Market Risk Premium}
\begin{itemize}
\item<1-> The \emph{market risk premium} or \emph{forward bias} $\pi (t,T)$
relates forward and expected spot prices.
\item<2-> It is defined as the difference, calculated at time $t$, between
the forward $F(t,T)$ at time $t$ with delivery at $T$ and expected
spot price:
\begin{equation}\label{forward risk premium}
\pi(t,T)= F(t,T)-\EX^\prb[S(T)|\mathcal F_t].
\end{equation}
Here $\EX^\prb$ is the expectation operator, under the
historical measure $\prb$, with information up until time $t$ and
$S(T)$ is the spot price at time $T$.
\end{itemize}

\subsection{Black's Formula}

% frametitle
{Simple Futures Market Model}
\begin{itemize}
\item<1-> We assume for the futures dynamics
$$
df(t) = \mu f(t) dt + \sigma f(t) dW(t).
$$
\item<2-> But since it costs nothing to enter a futures position, we can constantly change our futures portfolio.
Thus financing for buying a traded asset is not relevant.
\end{itemize}

% frametitle
{Black's Formula}

\begin{itemize}
\item<1-> We use the usual notation -
strike $K$, expiry $T$, $\tau=T-t$ for time to maturity as in the spot case, and write $\Phi$ for the
standard normal distribution function.
\item<2->
The arbitrage price $C$ of a European futures call option is
\begin{equation}\label{Black}
C(t)= e^{-r\tau} (f \Phi(d_1 (f,\tau)) - K \Phi(d_2 (f,\tau))),
\end{equation}
where
$$
d_{1,2} (f,\tau) := \frac{\log (f/K) \pm \frac{1}{2} {\sigma}^2 \tau}{
\sigma \sqrt{\tau}}.
$$
\end{itemize}

% frametitle
{Black's Formula}
\begin{itemize}
\item<1->
Observe that the quantities $d_1$ and $d_2$ do not depend on the
interest rate $r$.
\item<2-> This is intuitively clear from the classical
Black approach: one sets up a replicating risk-free portfolio
consisting of a position in futures options and an offsetting
position in the underlying futures contract. The portfolio
requires no initial investment and therefore should not earn any
interest.
\end{itemize}

\section{Bessembinder - Lemon model}

% frametitle
{Bessembinder- Lemon Model specification}
\begin{itemize}
\item<1-> One-period model
\item<2-> Power companies are able to forecast demand in the immediate future with precision
\item<3-> $N_P$ identical producers; $N_R$ identical retailers that buy power in the wholesale market and sell it to final consumers at fixed unit price
\item<4-> $P_R$ fixed unit price that consumers pay
\item<5-> $Q_{R_i}$ an exogenous random variable that denotes the realized demand for retailer $i$
\end{itemize}

% frametitle
{The cost function}
\begin{itemize}
\item<1-> Each producer $i$ has cost function
$$
TC_i=F+\frac{a}{c}(Q_{P_i})^c,
$$
where $F$ are fixed costs, $Q_{P_i}$ is the output of producer $i$, and $c\geq 2$.
\item<2-> The cost function implies that the marginal production costs increase with output.
\item<3-> If $c>2$ marginal costs increase at an increasing rate with output.
\item<4-> Moreover, the distribution of power prices will be positively skewed even when the distribution of power demand is symmetric.
\end{itemize}

% frametitle
{Clearing prices}
\begin{itemize}
\item<1-> First, assume that forward prices are gives
\item<2-> Obtain optimal behaviour in the spot market
\item<3-> Work back and find optimal positions in the forward market.
\end{itemize}

% frametitle
{The wholesale spot market}
\begin{itemize}
\item<1-> Producers sell to retailers who in turn distribute to power consumers
\item<2-> $P_W$ denotes the wholesale spot price, $Q_{P_i}^W$ quantity sold by producer $i$ in the wholesale spot market, $Q_{P_i}^F$ quantity
that producer $i$ has agreed to deliver (purchase if negative) in the forward market at the fixed forward price $P_F$.
\item<3-> The ex-post profit of producer $i$ is given by
$$
\pi_{P_i}=P_W Q_{P_i}^W + P_FQ_{P_i}^F
-F-\frac{a}{c}(Q_{P_i})^c,$$
where each producer's physical production, $Q_{P_i}$, is the sum of its spot and forward sales $Q_{P_i}^W+Q_{P_i}^F$.
\end{itemize}

% frametitle
{The wholesale spot market}
\begin{itemize}
\item<1-> Retailers buy in the real-time wholesale market the difference between realised retail demand and their forward positions
\item<2-> $Q_{R_j}^F$ quantity sold (purchased if negative) forward by retailer $j$, $P_R$ fixed retail price per unit
\item<3-> The ex-post profit for each retailer is
$$
\pi_{R_j}=P_R Q_{R_j} + P_FQ_{R_j}^F - P_W (Q_{R_j}+Q_{R_j}^F),$$
\item<4-> The profit maximising quantity for producer $i$ is (FOC wrt $Q_{P_i}^W$)
$$Q_{P_i}^W=\left(\frac{P_W}{a}\right)^{x}-Q_{P_i}^F$$
with $x=1/(c-1)$
\end{itemize}

% frametitle
{The wholesale spot market}
\begin{itemize}
\item<1-> The equilibrium total retail demand is equal to total production and forward contracts are in zero net supply
\item<2-> Hence we must have that summing over all producers production must equal total demand from retailers
$$
N_P\left(\frac{P_W}{a}\right)^{x}=\sum_{i=1}^{N_R}Q_{R_i}^F
$$
\end{itemize}

% frametitle
{The wholesale spot market}
\begin{itemize}
\item<1-> Therefore the market-clearing wholesale price is
$$
P_W=a \left(\frac{Q^D}{N_P}\right)^{c-1},$$
where $Q^D=\sum_{j=1}^{N_R}Q_{R_j}$ is total system demand. We see that when $c>2$ an increase in demand has a disproportionate effect on power prices.
\item<2-> Each producers sale in the wholesale market is
$$
Q_{P_i}^W= \frac{Q^D}{N_P}-Q_{P_i}^F.
$$
\end{itemize}

% frametitle
{Demand for forward positions}
\begin{itemize}
\item<1-> Producers profit (with no forwards) is
$$
\rho_{P_i}=P_W\frac{Q^D}{N_P}-F-\frac{a}{c}\left(\frac{Q^D}{N_P}\right)^{c}.$$
\item<2-> Retailers profit (with no forwards) is
$$
\rho_{R_j}=P_RQ_{R_j}-P_WQ_{R_j}.
$$
\end{itemize}

% frametitle
{Mean-Variance Analysis for optimal forward position}
Assume that market players
$$
\max_{Q^F_{\{P_i,R_j\}}}\EX[\pi_{\{P_i,R_j\}}]-\frac{A}{2}\var[\pi_{\{P_i,R_j\}}]
$$
where, for example, producers have the profit function
$$
\pi_{P_i}=\rho_{P_i}+ P^FQ^F-P_WQ^F.
$$
FOCs imply
$$
Q^F_{\{P_i,R_j\}}= \frac{P^F-\EX[P_W]}{A\var[P_W]}+\frac{\Cov[\rho_{\{P_i,R_j\}},P_W]}{\var[P_W]}.
$$

% frametitle
{Mean-Variance Analysis for optimal forward position}
\begin{itemize}
\item<1-> The optimal forward position contains two components
\begin{itemize}
\item The first term reflects the position taken in response to the bias $P^F-\EX[P_W]$
\item The second term is the quantity sold or bought forward to minimize the variance of profits
\end{itemize}
\item<2-> Forward hedging can reduce risk precisely because the covariance term is generally non-zero.
\end{itemize}

% frametitle
{The equilibrium forward price}
\begin{itemize}
\item<1-> One can show that
$$
P_F=\EX[P_W]-\frac{N_P}{Nca^x}\left[cP_R\Cov[P_W^{x}, P_W]-\Cov[P_W^{x+1}, P_W]\right],
$$
where $N=(N_R+N_P)/A$ reflects the number of firms in the industry and the degree to which they are concerned with risk.
\item<2-> The forward price will be less than the expected wholesale price, if the first term in brackets, which reflects retail risk, is
larger than the second term, which reflects production cost risk.
%\item<3-> The equilibrium forward price will depend only on the risks borne by the industry as a whole: variability in retail revenue and in %production costs
\end{itemize}

\input{ForwardsPart2}
